---
title: "group_assignment"
authors:   
    - "s252641 - Bhavya Gudhka"   
    - "s215065 - Henrik Servais Janssen Knudsen"   
    - "s253705 - Lun Suša"   

format:
  html:     
    embed-resources: true 
editor: visual 
execute:   
  warning: false   
  message: false 
---

### US Arrests - Data description

The dataset contains statistics about the arrests made per 100,000 residents in 50 states of US in 1973. The arrests are categorised into assault, murder, and rape. It also includes urban population percentage for different states. We carry out PCA on this data to reduce its dimensions and identify some patterns.

# PCA

#### Libraries

```{r}
library("tidyverse")
library("broom")
library("cowplot")
```

#### Load data

```{r}
library(ggplot2)
library(tidyverse)
library(broom)
library(cowplot)

# Perform PCA
pca_fit <- USArrests |> 
  select(where(is.numeric)) |> # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data

# Extract the rotation matrix (loadings) for arrows
arrow_data <- pca_fit |>
  tidy(matrix = "rotation") |>
  pivot_wider(names_from = "PC", 
              names_prefix = "PC", 
              values_from = "value") |>
  mutate(PC1 = PC1 * 5,  # Scale arrows for visibility
         PC2 = PC2 * 5)

# Plot with arrows
pca_fit |>
  augment(USArrests) |> 
  ggplot(aes(.fittedPC1, .fittedPC2)) + 
  geom_point(size = 1.5, color = "#0072B2") +
  geom_text(aes(label = rownames(USArrests)), 
            size = 3, vjust = -0.5, hjust = 0.5) +
  # Add arrows for variable loadings
  geom_segment(data = arrow_data,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "#D55E00",
               linewidth = 1) +
  # Add labels for arrows
  geom_text(data = arrow_data,
            aes(x = PC1, y = PC2, label = column),
            color = "#D55E00",
            size = 4,
            fontface = "bold",
            vjust = -0.5) +
  # Expand axis limits to show all labels
  coord_cartesian(xlim = c(-4, 4), ylim = c(-3.5, 5)) +
  theme_half_open(12) + 
  background_grid() +
  labs(title = "PCA Biplot of USArrests dataset\nPeople arrested and their correlation to different US states",
       x = "PC1",
       y = "PC2")
```

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:4) +  # Changed to 1:4 since USArrests has 4 variables
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12) +
  labs(title = "Scree Plot: \nVariance Explained by Principal Components",
       x = "Principal Component",
       y = "Percent Variance Explained")
```

#### PCA Analysis
